noam_data_exploration_v1

### 2024-05-30 noam_data_exploration_v2.Rmd
Add files via upload

site_colours <- c("LTER 1" = "cornflowerblue",
               "LTER 2" = "darkblue",
               "LTER 3" = "yellowgreen",
               "LTER 4" = "forestgreen",
               "LTER 5" = "mediumorchid",
               "LTER 6" = "orange")
habitat_linetypes <- c("Fringing" = "solid",
                       "Backreef" = "longdash",
                       "Outer 10" = "twodash",
                       "Outer 17" = "dotted")
habitat_shapes <- c("Fringing" = 16,
                     "Backreef" = 17,
                     "Outer 10" = 15,
                     "Outer 17" = 18)

@@

# WARNING - Before you run this, know that it will take a long time
set.seed(1032) 
ord2 <- metaMDS(species_matrix, dist	= "bray",	trymax	= 100,	k	= 2) # looks like it automatically wisconsin transforms now?
ord2 # stress = 0.22
stressplot(ord2)

@@

pro <- procrustes(ord2, ord3) # visually compare w procrustes rotation

plot(pro, cex=1.5)

@@

cent <- aggregate(cbind(NMDS1, NMDS2) ~ Stratum, data = site.scores, FUN = mean) # compute centroids for NMDS1 and NMDS2 
cent$NMDS1 <- as.numeric(cent$NMDS1) # read in as characters for some reason
cent$NMDS2 <- as.numeric(cent$NMDS2)
# Need to split stratum up for plotting later
cent<- 
  cent %>%
    separate(Stratum, into = c("Year", "Site", "Habitat"), sep = "-")
cent$Site_habitat <- paste(cent$Site, cent$Habitat)
# Add back in stratum column
cent$Stratum <-  aggregate(cbind(NMDS1, NMDS2) ~ Stratum, data = site.scores, FUN = mean)$Stratum
# A lot of extra spaces

cent <- cent %>%
  mutate(across(everything(), str_trim))
cent$Year <- as.factor(cent$Year)
cent$Habitat <- as.factor(cent$Habitat)
cent$Site <- as.factor(cent$Site)
cent$Site_habitat <- as.factor(cent$Site_habitat)

species.scores <- as.data.frame(scores(ord3)$species)  # extract the species scores and convert to a data.frame
species.scores$Species <- rownames(species.scores)

ggplot() + 
  geom_point(data= site.scores, aes(x=NMDS1,y=NMDS2, shape = Habitat, colour = Site),
             alpha = 0.25, size=2) + # add the point markers
#  geom_text(data= species.scores,aes(x=NMDS1,y=NMDS2, label = Species), size = 4) +  # add the species labels

@@

```{r CV, fig.height = 10, fig.width = 10}
# coefficient of variation
CV<-function(x){
  return(sd(x,na.rm=T)/mean(x,na.rm=T))
}

@@

diversity_stability <- alpha_diversity %>%
  dplyr::group_by(Site, Habitat, Site_habitat) %>%
  summarise(Richness_mean = mean(Richness), 
            Richness_se = sd(Richness)/sqrt(n()),
            Richness_cv = CV(Richness),
            Shannon_mean = mean(Shannon), 
            Shannon_se = sd(Shannon)/sqrt(n()),
            Shannon_cv = CV(Shannon),
            Cover_mean = mean(Cover),
            Cover_se = sd(Cover)/sqrt(n()),
            Cover_cv = CV(Cover))

diversity_mod_1 <- glmmTMB(Cover_cv ~ Richness_mean, family = Gamma(link = "log"), data = diversity_stability)
summary(diversity_mod_1)
plot(allEffects(diversity_mod_1)) # richness reduces cv (i.e. enhances diversity!)

@@

diversity_mod_5 <- glmmTMB(Cover_cv ~ Site, family = Gamma(link = "log"), data = diversity_stability)
summary(diversity_mod_5)
Anova(diversity_mod_5) # first non-significant model! Variance in cover not driven by site
# Also makes sense given nmds plots above
# Maybe site as a random effect similar to Cline and Allgeier 2022?

@@

AICc(diversity_mod_1, diversity_mod_2, diversity_mod_4, diversity_mod_6) # Model with just Shannon diversity best

@@

(shannon_cv_plot = 
  ggplot(data = diversity_stability) +
  geom_point(aes(x = Shannon_mean, y = Cover_cv, colour = Site, shape = Habitat), size = 8) + 
   scale_colour_manual(values = site_colours) +
  scale_shape_manual(values = habitat_shapes) +
  theme_bw() + 
  theme(axis.title.x = element_text(size=18), 
        axis.title.y = element_text(size=18), 
        panel.background = element_blank(), 
        panel.grid.major = element_blank(),  #remove major-grid labels
        panel.grid.minor = element_blank(),  #remove minor-grid labels
        plot.background = element_blank(),
        legend.text = element_text(size = rel(1)), # increase legend text
        legend.title = element_text(size = rel(1)),
        legend.position = 'bottom', # move legend to bottom
        legend.box = 'horizontal', # make legend horizontal
        legend.box.background = element_rect(colour = "black"))  # add box around legend
         )

# maybe some heteroskedasticity?
# just eyeballing but it looks like "stability" is highest in LTER 2 Fringe, which also has low diversity
# Possibility for context-dependence. Alternative mechanisms drive stability?
# Functional form vs. portfolio effects?

# Stopping for now
# GOALS: test portfolio effects, look at turnover and temporal beta diversity stats, ways to test beta dispersion

@@

>>>>

site_colours <- c("LTER 1" = "cornflowerblue",
               "LTER 2" = "darkblue",
               "LTER 3" = "yellowgreen",
               "LTER 4" = "forestgreen",
               "LTER 5" = "mediumorchid",
               "LTER 6" = "orange")
habitat_linetypes <- c("Fringing" = "solid",
                       "Backreef" = "longdash",
                       "Outer 10" = "twodash",
                       "Outer 17" = "dotted")
habitat_shapes <- c("Fringing" = 16,
                     "Backreef" = 17,
                     "Outer 10" = 1,
                     "Outer 17" = 2)

@@

# identifying species w/ > 10% cover for plotting later
# NOTE: Used ChatGPT to generate these next few lines of code but it looked good so I went with it
columns_to_keep <- apply(species_matrix, 2, function(x) any(x >= 10))
abundant_species_matrix <- as.data.frame(species_matrix)[ , c(TRUE, columns_to_keep)]
abundant_species_names <- colnames(abundant_species_matrix)

# WARNING - Before you run this, know that it will take a long time
set.seed(1032) 
# ord2 <- metaMDS(species_matrix, dist	= "bray",	trymax	= 100,	k	= 2) # looks like it automatically wisconsin transforms now?
# ord2 # stress = 0.22
# stressplot(ord2)

@@

# pro <- procrustes(ord2, ord3) # visually compare w procrustes rotation

# plot(pro, cex=1.5)

@@

cent <- aggregate(cbind(NMDS1, NMDS2) ~ Stratum, data = site.scores, FUN = mean) # compute centroids for NMDS1 and NMDS2

# Need to split stratum up for plotting later
cent<- 
  cent %>%
    separate(Stratum, into = c("Year", "Site", "Habitat"), sep = "-")
cent$Site_habitat <- paste(cent$Site, cent$Habitat)
# Add back in stratum column
cent$Stratum <-  aggregate(cbind(NMDS1, NMDS2) ~ Stratum, data = site.scores, FUN = mean)$Stratum
# A lot of extra spaces

cent <- cent %>%
  mutate(across(everything(), str_trim))
cent$Year <- as.factor(cent$Year)
cent$Habitat <- as.factor(cent$Habitat)
cent$Site <- as.factor(cent$Site)
cent$Site_habitat <- as.factor(cent$Site_habitat)
cent$NMDS1 <- as.numeric(cent$NMDS1) # read in as characters for some reason
cent$NMDS2 <- as.numeric(cent$NMDS2)
species.scores <- as.data.frame(scores(ord3)$species)  # extract the species scores and convert to a data.frame
species.scores$Species <- rownames(species.scores)

# subset only the species w/ > 10% cover
abundant.species.scores <- species.scores[species.scores$Species %in% abundant_species_names,]

ggplot() + 
  geom_point(data= site.scores, aes(x=NMDS1,y=NMDS2, shape = Habitat, colour = Site),
             alpha = 0.25, size=2) + # add the point markers
  geom_text(data= abundant.species.scores,aes(x=NMDS1,y=NMDS2, label = Species), size = 4) +  # add the species labels

@@

```{r CV, fig.height = 8, fig.width = 10}
# coefficient of variation
CV<-function(x){
  return(sd(x,na.rm=T)/mean(x,na.rm=T))
}

@@

diversity_stability <- alpha_diversity %>%
  dplyr::group_by(Site, Habitat, Site_habitat, Transect) %>%
  summarise(Richness_mean = mean(Richness), 
            Richness_se = sd(Richness)/sqrt(n()),
            Richness_cv = CV(Richness),
            Shannon_mean = mean(Shannon), 
            Shannon_se = sd(Shannon)/sqrt(n()),
            Shannon_cv = CV(Shannon),
            Cover_mean = mean(Cover),
            Cover_se = sd(Cover)/sqrt(n()),
            Cover_cv = CV(Cover))

diversity_mod_1 <- glmmTMB(Cover_cv ~ Richness_mean, family = Gamma(link = "log"), data = diversity_stability)
summary(diversity_mod_1)
plot(allEffects(diversity_mod_1)) # richness reduces cv (i.e. enhances stability!)

@@

diversity_mod_5 <- glmmTMB(Cover_cv ~ Site, family = Gamma(link = "log"), data = diversity_stability)
summary(diversity_mod_5)
Anova(diversity_mod_5)

@@

diversity_mod_7 <- glmmTMB(Cover_cv ~ Shannon_mean*Site + (1|Site_habitat), family = Gamma(link = "log"), data = diversity_stability)
summary(diversity_mod_7)
Anova(diversity_mod_7)

diversity_mod_universal <- glmmTMB(Cover_cv ~ Shannon_mean*Site*Habitat + (1|Site_habitat), family = Gamma(link = "log"), data = diversity_stability, na.action = na.pass)
diversity_dredge <- dredge(diversity_mod_universal)
model.sel(diversity_dredge)

# best model based on dredge:
diversity_best <-  glmmTMB(Cover_cv ~ Shannon_mean*Site + Site*Habitat + (1|Site_habitat), family = Gamma(link = "log"), data = diversity_stability)
summary(diversity_best) # very low variance in random effect, which is great
Anova(diversity_best) # Interactions all significant
plot(dp <- simulateResiduals(diversity_best)) # some outliers, but otherwise good model fit. I am fine with this.

@@

(shannon_cv_plot = 
  ggplot(data = diversity_stability) +
  geom_point(aes(x = Shannon_mean, y = Cover_cv, colour = Site, shape = Habitat), size = 8, stroke = 2) + 
   scale_colour_manual(values = site_colours) +
   scale_shape_manual(values = habitat_shapes) +
  facet_wrap(~Site) +
  theme_bw() + 
  theme(axis.title.x = element_text(size=18), 
        axis.title.y = element_text(size=18), 
        panel.background = element_blank(), 
        panel.grid.major = element_blank(),  #remove major-grid labels
        panel.grid.minor = element_blank(),  #remove minor-grid labels
        plot.background = element_blank(),
        legend.text = element_text(size = rel(1)), # increase legend text
        legend.title = element_text(size = rel(1)),
        legend.position = 'bottom', # move legend to bottom
        legend.box = 'horizontal', # make legend horizontal
        legend.box.background = element_rect(colour = "black"))  # add box around legend
         )

# maybe some heteroskedasticity?
# just eyeballing but it looks like "stability" is highest in LTER 2 Fringe, which also has low diversity
# Functional form vs. portfolio effects?
# Either way looks sick as shit. Great clustering by habitat within site, and diversity trend is strong in LTER 3 4 5
# LTER2 the trend is reversed
# LTER 1 is mostly driven by habitat
# LTER 6 potentially a wash

# GOALS: test portfolio effects, look at turnover and temporal beta diversity stats, ways to test beta dispersion

@@

```{r variance vs mean diversity plot, fig.height = 8, fig.width = 10}
(diversity_mean_cv_plot = 
  ggplot(data = diversity_stability) +
  geom_point(aes(x = Shannon_mean, y = Shannon_cv, colour = Site, shape = Habitat), size = 8, stroke = 2) + 
   scale_colour_manual(values = site_colours) +
   scale_shape_manual(values = habitat_shapes) +
  facet_wrap(~Site) +
  theme_bw() + 
  theme(axis.title.x = element_text(size=18), 
        axis.title.y = element_text(size=18), 
        panel.background = element_blank(), 
        panel.grid.major = element_blank(),  #remove major-grid labels
        panel.grid.minor = element_blank(),  #remove minor-grid labels
        plot.background = element_blank(),
        legend.text = element_text(size = rel(1)), # increase legend text
        legend.title = element_text(size = rel(1)),
        legend.position = 'bottom', # move legend to bottom
        legend.box = 'horizontal', # make legend horizontal
        legend.box.background = element_rect(colour = "black"))  # add box around legend
         )
# More diverse regions are also those that fluctuate the least in their diversity
```

```{r general cover diversity, fig.height = 30, fig.width = 30}
(diversity_cover_plot = 
  ggplot(data = alpha_diversity) +
  geom_point(aes(x = Cover, y = Shannon, colour = Site, shape = Habitat), size = 8, stroke = 2) + 
   scale_colour_manual(values = site_colours) +
   scale_shape_manual(values = habitat_shapes) +
  facet_wrap(~Year) +
  theme_bw() + 
  theme(axis.title.x = element_text(size=18), 
        axis.title.y = element_text(size=18), 
        panel.background = element_blank(), 
        panel.grid.major = element_blank(),  #remove major-grid labels
        panel.grid.minor = element_blank(),  #remove minor-grid labels
        plot.background = element_blank(),
        legend.text = element_text(size = rel(1)), # increase legend text
        legend.title = element_text(size = rel(1)),
        legend.position = 'bottom', # move legend to bottom
        legend.box = 'horizontal', # make legend horizontal
        legend.box.background = element_rect(colour = "black"))  # add box around legend
         ) # ABORT ABORT ABORT
```
